### 2024-08-29 teaching/math240f24/draft.tex
Teaching

\datestamp{1}{}
A {\it set} is a collection of distinct objects, called its {\it elements} or its {\it members}. If
$x$ is a member of set $A$, then we write $x\in A$, and if $x$ is not an element of $A$,
then we write $x\notin A$.
Sets can be written by listing out its elements. For example,
$$\{1,4,7,10,\sqrt{782}\}$$
and
$$\bigl\{\{1,2\}, \pi, \{4\}\bigr\}$$
are both sets (the second example shows that sets can themselves contain other sets). The order of the elements
is not important, and duplicate elements are ignored (so $\{1,2\} = \{2,1\} = \{1,1,2\}$).
The notation using $\{$ and $\}$
is useful for defining small, concrete examples, but expressing large sets can become
very cumbersome. The first way one can describe larger sets is to use the $\ldots$ symbol and the power of
suggestion. For instance, anyone faced with the notation
$$A = \{1,3,5,7,9,\ldots\}$$
can quickly guess that this set is supposed to contain all the positive odd integers. We can also use $\ldots$
to define finite sets. Most Canadians will be able to tell you that the set
$$\{\hbox{\rm Alberta}, \hbox{\rm British Columbia}, \ldots, \hbox{\rm Yukon}\}$$
of provinces and territories contains $13$ elements.
But this notation inherently produces some ambiguity. For example, since the sequence of positive palindromic
binary numbers starts $1,3,5,7,9,15,17,21,27,\ldots$,
we are left with some doubt as to whether the set $A$ above should be the set of odd numbers or the set of
palindromic binary numbers.

@@

\proclaim Proposition \advthm. If $A$ is finite, then
$|2^A| = 2^{|A|}$.

@@

Here are some examples and non-examples.
\medskip
\item{i)} The function $f:\RR\to\RR$ given by $f(x) = x^2$ is
$$f = \bigl\{ (a,b)\in \RR^2 : b = a^2 \bigl\},$$
when written in set-builder notation.
\smallskip
\item{ii)} On the other hand, the set
$$g = \bigl\{ (a,b)\in \RR^2 : a = b^2 \bigl\}$$
is not a function, since $(1,1)$ and $(1,-1)$ are both in $g$. Furthermore,
there is no element of $g$ with $-1$ as its first coordinate.
\smallskip
\item{iii)} If $X = \{1,4,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
$$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
is a function.
\smallskip
\item{iv)} The set
$$h = \bigl\{ (x,y)\in \RR^2 : y = 1/x \bigl\}$$
is not a function, since there is no $y$ such that $(0,y)\in h$. However
if we amend the domain and consider
$$h = \bigl\{ (x,y)\in \bigl(\RR\setminus \{0\}\bigr)\times \RR : y = 1/x \bigl\},$$
then in fact, $h:\RR\setminus \{0\} \to \RR$ is a function.
\medskip
\datestamp{8}{}
The {\it range} or {\it image} of a function $f:A\to B$ is
$$f(A) = \bigl\{b\in B : \hbox{there exists}\ a\in A\ \hbox{such that}\ b = f(a)\bigr\}.$$
These are all the values $f$ actually outputs. For instance, if we let $f : \ZZ\to \NN$
be given by $f(n) = n^2$, then
$$\eqalign{
f(\ZZ) &= \{\ldots, f(-2), f(-1), f(0) , f(1), f(2),\ldots\} \cr
&= \{\ldots, 9, 4, 1, 0 1, 4, 9,\ldots\} \cr
&= \{0, 1, 4, 9, 16, 25, \ldots\}. \cr
}$$
A function $f:A\to B$ is called {\it surjective} or {\it onto} if $f(A) = B$, that is,
for every $b\in B$ there exists some $a\in A$ such that $f(a) = b$. An example of a
surjective function is $f:\QQ\to \QQ$ sending $x\mapsto x/2$. This is because for any $q\in \QQ$,
we can set $r = 2q$, and
$$f(r) = {r\over 2} = {2q\over 2} = q.$$
A function $f:A\to B$ is called {\it injective} or {\it one-to-one} if for all $a_1,a_2\in A$
with $a_1 \ne a_2$, we also have $f(a_1)\ne f(a_2)$. Equivalently, $f$ is injective if
$f(a_1) = f(a_2)$ implies that $a_1 = a_2$ for all $a_1, a_2 \in A$. For instance, the function
$f:\ZZ\to \NN$ that sends $n\mapsto n^2$ is not injective since $f(-1) = f(1)$, but $-1 \ne 1$.
On the other hand, if we modify the domain, considering $f:\NN\to \NN$ sending $n\mapsto n^2$,
then now $f$ is injective, since if $f(m) = f(n)$, then $m^2 = n^2$, and there is only one positive
integer that squares to any given integer, so $m=n$.

@@

\medskip\boldlabel The pigeonhole principle. We now take a brief pause to introduce one of
the most fundamental laws in discrete mathematics. It is called the {\it pigeonhole principle},
because one formulation of it is, {\sl If $n$ pigeons nest in $n-1$ holes, then there is at least
one hole that contains more than one pigeon.}
First we'll prove a more general statement.

@@

This statement can be summed up in one sentence: {\sl The maximum is at least the average.}

A function $f$ is called {\it bijective} (or a {\it one-to-one correspondence})
if it is injective and surjective. Bijections are important because of the following proposition.

@@

\proof Suppose $|A| = |B| = n$. Then choose an ordering $a_1, \ldots, a_n$ of $A$ and an ordering
$b_1, \ldots, b_n$ of $B$. Let $f(a_i) = b_i$ for all $1\le i\le n$. By construction, this is a bijection.
On the other hand, suppose $|A|\ne |B|$. If $A<B$, then $f$ cannot be surjective, since the image of
$f$ has size at most $|A| < |B|$ (at least one element of $B$ must be missed). If $A>B$, then since
$f(A)$ must

>>>>

\datestamp{1}{}
A {\it set} is a collection of distinct objects, called its {\it elements} or its {\it members}. If
$x$ is a member of set $A$, then we write $x\in A$, and if $x$ is not an element of $A$,
then we write $x\notin A$.
Sets can be written by listing out its elements. For example,
$$\bigl\{1,4,7,10,\sqrt{782}\bigr\}$$
and
$$\bigl\{\{1,2\}, \pi, \{4\}\bigr\}$$
are both sets (the second example shows that sets can themselves contain other sets). The order of the elements
is not important, and duplicate elements are ignored (so $\{1,2\} = \{2,1\} = \{1,1,2\}$).
The notation using $\{$ and $\}$
is useful for defining small, concrete examples, but expressing large sets can become
very cumbersome. The first way one can describe larger sets is to use the $\ldots$ symbol and the power of
suggestion. For instance, anyone faced with the notation
$$A = \{1,3,5,7,9,\ldots\}$$
can quickly guess that this set is supposed to contain all the positive odd integers. We can also use $\ldots$
to define finite sets. Most Canadians will be able to tell you that the set
$$\{\hbox{\rm Alberta}, \hbox{\rm British Columbia}, \ldots, \hbox{\rm Yukon}\}$$
of provinces and territories contains $13$ elements.
But this notation inherently produces some ambiguity. For example, since the sequence of positive palindromic
binary numbers starts $1,3,5,7,9,15,17,21,27,\ldots$,
we are left with some doubt as to whether the set $A$ above should be the set of odd numbers or the set of
palindromic binary numbers.

@@

\edef\propcardpowerset{\the\thmcount}
\proclaim Proposition \advthm. If $A$ is finite, then
$|2^A| = 2^{|A|}$.

@@

Here are some examples and non-examples.
\medskip
\item{i)} The function $f:\RR\to\RR$ given by $f(x) = x^2$ is
$$f = \bigl\{ (a,b)\in \RR^2 : b = a^2 \bigl\},$$
when written in set-builder notation.
\smallskip
\item{ii)} On the other hand, the set
$$g = \bigl\{ (a,b)\in \RR^2 : a = b^2 \bigl\}$$
is not a function, since $(1,1)$ and $(1,-1)$ are both in $g$. Furthermore,
there is no element of $g$ with $-1$ as its first coordinate.
\smallskip
\item{iii)} If $X = \{1,4,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
$$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
is a function.
\smallskip
\item{iv)} The set
$$h = \bigl\{ (x,y)\in \RR^2 : y = 1/x \bigl\}$$
is not a function, since there is no $y$ such that $(0,y)\in h$. However
if we amend the domain and consider
$$h = \bigl\{ (x,y)\in \bigl(\RR\setminus \{0\}\bigr)\times \RR : y = 1/x \bigl\},$$
then in fact, $h:\RR\setminus \{0\} \to \RR$ is a function.
\medskip

\medskip\boldlabel\datestamp{8}{} Injective and surjective functions.
The {\it range} or {\it image} of a function $f:A\to B$ is
$$f(A) = \bigl\{b\in B : \hbox{there exists}\ a\in A\ \hbox{such that}\ b = f(a)\bigr\}.$$
These are all the values $f$ actually outputs. For instance, if we let $f : \ZZ\to \NN$
be given by $f(n) = n^2$, then
$$\eqalign{
f(\ZZ) &= \{\ldots, f(-2), f(-1), f(0) , f(1), f(2),\ldots\} \cr
&= \{\ldots, 9, 4, 1, 0 1, 4, 9,\ldots\} \cr
&= \{0, 1, 4, 9, 16, 25, \ldots\}. \cr
}$$
A function $f:A\to B$ is called {\it surjective} or {\it onto} if $f(A) = B$, that is,
for every $b\in B$ there exists some $a\in A$ such that $f(a) = b$. An example of a
surjective function is $f:\QQ\to \QQ$ sending $x\mapsto x/2$. This is because for any $q\in \QQ$,
we can set $r = 2q$, and
$$f(r) = {r\over 2} = {2q\over 2} = q.$$
A function $f:A\to B$ is called {\it injective} or {\it one-to-one} if for all $a_1,a_2\in A$
with $a_1 \ne a_2$, we also have $f(a_1)\ne f(a_2)$. Equivalently, $f$ is injective if
$f(a_1) = f(a_2)$ implies that $a_1 = a_2$ for all $a_1, a_2 \in A$. For instance, the function
$f:\ZZ\to \NN$ that sends $n\mapsto n^2$ is not injective since $f(-1) = f(1)$, but $-1 \ne 1$.
On the other hand, if we modify the domain, considering $f:\NN\to \NN$ sending $n\mapsto n^2$,
then now $f$ is injective, since if $f(m) = f(n)$, then $m^2 = n^2$, and there is only one positive
integer that squares to any given integer, so $m=n$.

@@

\medskip\boldlabel The pigeonhole principle. We now take a brief pause to introduce one of
the most fundamental laws in discrete mathematics, called the pigeonhole principle.
We begin with the following intuitive theorem.

@@

This theorem can be summed up in one sentence: {\sl The maximum is at least the average.}
Don't underestimate this theorem even though its proof was a one-liner! It is often used to prove
highly nontrivial results.
(As an exercise, prove the similar statement: {\sl The minimum is at most the average.})
From here we are now equipped to prove (a general version of) the pigeonhole principle.

\edef\thmgeneralpigeonhole{\the\thmcount}
\proclaim Theorem \advthm. Let $A$ and $B$ be finite sets with $|A| = m$ and $|B|=n$.
For every function $f:A\to B$, then there is some $b\in B$ such that there are at least
$\lceil m/n\rceil$ elements $a\in A$ with $f(a) = b$.

\proof For $1\le i \le n$, let $r_i$ be the number of $a\in A$
such that $f(a) = b_i$. This is a sequence that adds up to $m$, since every $a$ in $A$ maps to
exactly one element of $B$. So the sequence has average $m/n$, and by the previous theorem,
there must be some $j$ such that $r_j \ge m/n$. But the $r_i$ are all actually integers (since
cardinalities of finite sets are integers), meaning that $r_j \ge \lceil m/n \rceil$. Letting
$b = b_j$ completes the proof.\slug

The reason this is called the pigeonhole principle is because of the following special case.

\parenproclaim Corollary {\advthm} (Pigeonhole principle). Let $n\ge 2$. If $n$ pigeons nest in $n-1$ holes,
there is at least one hole that contains at least two pigeons.

\proof Let $A$ be the set of pigeons and $B$ the set of pigeonholes. Let $f$ be any function sending
the set of pigeons to the set of holes. By the previous theorem, there is some hole with at least
$\lceil n/(n-1)\rceil = 2$ pigeons in it. (This is because $1 < n/(n-1) < 2$ for all integers $n\ge 2$.)\slug

\medskip\boldlabel Bijections.
A function $f$ is called {\it bijective} (or a {\it bijection}, or a {\it one-to-one correspondence})
if it is injective and surjective. Bijections are important because of the following proposition.

@@

\proof Suppose $|A| = |B| = n$. Then choose an ordering $a_1, \ldots, a_n$ of $A$ and an ordering
$b_1, \ldots, b_n$ of $B$. Let $f(a_i) = b_i$ for all $1\le i\le n$. By construction, this is a bijection,
proving one direction of (i).

On the other hand, suppose $|A|\ne |B|$ (so we prove this direction by contraposition).
If $A<B$, then $f$ cannot be surjective, since the image of
$f$ has size at most $|A| < |B|$ (at least one element of $B$ must be missed). If $A>B$, then
$|A|/|B| > 1$, so by Theorem~{\thmgeneralpigeonhole}, there is some element $b\in B$
such that the number of $a\in A$ mapping to $b$ is at least $\lceil |A|/|B| \rceil \ge 2$.
This means that $f$ is not injective. We have proved part (i).

To prove part (ii), let $|A| = |B|$ and let $f:A\to B$. First we assume that $f$ is injective.
We enumerate $A = \{a_1, a_2, \ldots, a_n\}$. Then
$$f(A) = \bigl\{ f(a_1), f(a_2), \ldots, f(a_n)\bigr\} \subseteq B.$$
All of the $f(a_i)$ are distinct, since if $f(a_i) = f(a_j)$, then $a_i = a_j$. So
$\bigl| f(A)\bigr| = |A| = n$, and $f(A)$ is a size $n$ subset of $B$, which also has size $n$.
Hence $f(A) = B$; that is, $f$ is surjective.

Lastly, suppose $f$ is not injective (again we are using contraposition). So there
are $a_i$ and $a_j$ such that $a_i\ne a_j$ but $f(a_i) = f(a_j)$. So
$$\bigl|f(A)\bigr| = \bigl|\bigl\{ f(a_1), f(a_2), \ldots, f(a_n)\bigr\}\bigr| < n = |B|,$$
so $f(A) \ne B$ and $f$ is not surjective.\slug

Item (i) of the previous proposition should be entirely intuitive, especially if we use the alternative
nomenclature ``one-to-one correspondence'' instead of ``bijection.'' (In fact, we already implicitly
used (i) in these notes, in the proof of Proposition~{\propcardpowerset}.) Item (ii) is perhaps not
as immediate, but should become clear if you work it out with a picture.

\medskip\boldlabel Bijections.
A function $f:A\to B$ is called {\it invertible} if there exists $g:B\to A$ such that
\medskip
\item{i)} for all $b\in B$, $f\bigl(g(b)\bigr) = b$; and
\smallskip
\item{ii)} for all $a\in A$, $g\bigl(f(a)\bigr) = a$.
\medskip
If $g$ exists, it can be shown that $g$ must be unique, so we write $g = f^{-1}$ and speak of
{\it the} inverse of $f$.

\proclaim Proposition \advthm. Let $f:A\to B$ be a function. Then $f$ is invertible
if and only if $f$ is bijective.

\proof First we assume that $f$ is invertible. So there exists an inverse $g$ of $f$. For each $b\in B$,
setting $a = g(b)$ we have
$$f(a) = f\bigl(g(b)\bigr) = b.$$
This proves that $f$ is surjective. To show that $f$ is injective, suppose that $f(a_1) = f(a_2)$.
By applying $g$ on both sides, we have $g\bigl(f(a_1)\bigr) = g\bigl(f(a_2)\bigr)$, whence
$a_1 = a_2$, by definition of $g$.

Now assume that $f$ is bijective. We need to define $g:B\to A$. Well, gven any $b\in B$, there is
some $a$ such that $f(a) = b$, from surjectivity of $f$, and this $a$ is unique, since $f$ is
injective. So set $g(b) = a$ (and repeat this process for every $b\in B$). We have
$f\bigl(g(b)\bigr) = f(a) = b$, and for every $a\in A$, by definition of $g$ the
element $g\bigl(f(a)\bigr)$ is the unique element in $A$ that gets brought to $f(a)$ by $f$,
has to be $a$ itself.\slug

Sometimes to prove that two sets have the same cardinality, it is easier to prove that there
exists a bijection (as we already saw in the example of Proposition~{\propcardpowerset}),
and sometimes to prove that a function is a bijection, it is easier to show that it has an inverse,
rather than messing around with the definitions of injective and surjective.
Here's an example.

\proclaim Proposition. Let $X$ be a finite nonempty set. Let $E$ be the set of all subsets of $X$ with
even cardinality, and let $F$ be the set of all subsets of $X$ with odd cardinality.
Then $|E| = |F|$.

\proof We shall construct a function $f:E\to F$. Fix one specific $x\in X$;
we can do this because $X\ne \emptyset$. Now, for all $A\in E$, let
$$f(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
Note that since $|A|$ is even for all $A\in E$, the cardinality of $f(A)$ is odd (in the first
case it is $|A|-1$ and in the second case it is $|A|+1$. This shows that $f$ is indeed a function
with codomain $F$. Now to prove $|E| = |F|$ we will show that $f$ is bijective, which we
do by showing that $f$ is an inverse (as an exercise, you might try to prove bijectivity from
the definitions of injective and surjective).

We define $g : F\to E$ by
$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
for any $A\in E$,
$$\eqalign{
g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
&= \cases{A, & if $x\in A$;\cr
A, & if $x\notin A$\cr} \cr
&= A.\cr
}$$
The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug

\advsect Cardinality

\datestamp{9}{}
Earlier, we defined the cardinality of a set to be the number of elements it contains. What, then,
is the cardinality of $\NN$? How about $\RR$? You might say $\infty$, but this is not a number
(at least, it's not an element of $\NN$, the way all cardinalities of finite sets are).
So perhaps we should amend our question to the following: {\sl When do infinite sets have the same size?}
Our experience with functions leads us to the answer: {\sl When there exists a bijection between
them}. We shall say that $A$ and $B$ are {\it equipotent} (or {\it equinumerous}, or {\it have
the same cardinality}) if there exists a bijection between $A$ and $B$. In this case
we write $|A| = |B|$.

As an example, the sets $\NN$ and $\NN\setminus\{0\}$ are equipotent, since $f$
given by $n\mapsto n+1$ is a bijection $\NN\to\NN\setminus \{0\}$. (Check that
$f^{-1}(m) = m-1$ is its inverse.)

It is even possible to remove an infinite number of elements from $\NN$ and end up with
something still equipotent with $\NN$. To see this, let $E$ be the set of nonnegative even integers,
and consider the function $f :\NN\to E$ sending $n\mapsto 2n$. This is injective because if
$2m = 2n$, then dividing out by $2$ on both sides yields $m=n$. It is surjective because if
$n\in E$, then $n=2k$ for some $k\in\NN$, by definition, and $f(k) = 2k = n$.
